---
title: "Lab02"
author: "Elijah Boulton"
date: "January 24, 2019"
output: 
  html_document:
    toc: yes
    toc_float: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Task 1

The working directory

```{r}
getwd()
```

# Task 2

First six lines of data

```{r}

epagas = read.csv("EPAGAS.csv")
head(epagas)
```

# Task 3

Calculate z values

```{r}
mpg = epagas$MPG
z = (mpg - mean(mpg))/sd(mpg)
#print(mean(z)) #printed to verify results
#print(sd(z)) #printed to verify results
```
Possible outliers

```{r}
print(mpg[abs(z) >= 2 & abs(z) <= 3])
```

Definite outliers

```{r}
print(mpg[abs(z) > 3])
```

Dotplot of MPG with outliers in red and possible outliers in blue

```{r}
library(lattice)

mycolor= ifelse(abs(z) > 3, "Red",
         ifelse(abs(z) >= 2 & abs(z) <= 3, "Blue", "Black"))

dotplot(mpg, col = mycolor, main = "MPG with outliers")
```

# Task 4

Boxplot of the data in mpg

```{r}
boxplot(mpg, main = "Miles Per Gallon", notch = TRUE, horizontal = TRUE,
        col = "Black")
```

According to Chebyshev's theorem, approximately 75% of the data should be within two standard deviations of the mean.

The proportion of data that actually falls within two standard deviations of the mean

```{r}
temp = mpg[mpg >= mean(mpg) - 2*sd(mpg) &
             mpg <= mean(mpg) + 2*sd(mpg)]

percentData = length(temp)/length(mpg)

print(percentData)

```

Chebyshev's rule does not agree with the data seeing as it underestimates the proportion of the data within two standard deviations of the mean by 21%.

According to the empirical rule, approximately 95% of the data should be within two standard deviations of the mean.

The empirical rule slightly underestimates the cohesiveness of the data, but still fits pretty well.

The empirical rule is valid because it fits the data quite well, despite the relatively small sample size of only 100 measurements.